\documentclass[a4paper,12pt]{article}

\usepackage[backend=bibtex]{biblatex}
\usepackage{geometry}
\usepackage{titling}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage[hidelinks]{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{forest}
\usepackage{tikz-qtree}
\usepackage{setspace}

\addbibresource{ref.bib}

\titleformat{\section}
{\Huge}
{}
{0em}
{}[\titlerule]
\geometry{a4paper,total={170mm,257mm},left=25mm,right=25mm,}

\author{Lucas Standen}
\title{WORKING TITLE: Why FOSS software is preferred in the 
development and privacy space?}


\begin{document}
\maketitle

\newpage

\section{Using this document}
This document is written using the {\LaTeX} text compiler. The compiler has set up clickable links,
clickable references and a clickable table of contents, so please use these to your advantage. 
The Tex source and Bib Tex bibliography is available for all at 
\url{https://github.com/standenboy/epq/}.

\tableofcontents
\newpage

\setlength{\parskip}{1em}

{\setlength{\parindent}{0cm}

\section{A brief introduction}

\section{Used language in this paper}
Throughout this paper I will use language specific to the field of computer science, and as such
it makes sense to provide a brief overview for those who don't know what specific terms mean.

\begin{description}
	\item[Licenses] In this setting a license is a legal document that is distributed with
		almost all modern software, which describes how someone can use a piece of software
	\item[Free Software] This term refers to software under specific licenses, making them 
		free for the user to use (free as in freedom, not the monetary cost). This will
		be covered further in the next section.
	\item[Open Source] This term refers to a piece of software, where the original code for it
		is publicly available. This too will be covered further in the next section.
	\item[FOSS] An acronym for "Free and Open Source Software".
\end{description}

\section{What is Free Software?}
The Free Software movement is one that has been active for over 40 years \cite{GNUmaifesto}, it has
created some of the most important tools in computing that are used by billions on a daily basis. 
It is so engraved in our lives, yet so few even know what the term means; In a simple note, it is
software for a computer, phone or other device that can be used without violating the users 
freedom.

The definition of what counts Free Software and what is software freedom can vary depending on who 
you ask, but it was originally written that software that allows the following freedoms is 
Free Software:

\begin{description}
	\item[0] The freedom to run a program for any purpose
	\item[1] The freedom to study how a program works, and modify it to your needs
	\item[2] The freedom to redistribute a piece of software
	\item[3] The freedom to redistribute a edited version of software publicly
\end{description}
\textit{These freedoms were written by Richard Stallman\cite{FOSSdef} who is ever 
	important in this space.}

It is important that one does not confuse Free Software with software that is monetarily free, 
this is known as Freeware. Free Software defends the users rights to use and modify software and
is not focused on its cost.

One should also note the differences between Free Software and Open Source software. In Open Source
software, like Free Software, the original code for a program is available to anyone, however
in Open Source, this is to better the projects development and usability, whereas in Free Software
it is to better the users freedom. They both use the same methods to achieve differing goals; this
often leads them to be commonly used together, as the benefits a user gets from Free Software is 
much the same in Open Source software, and vice versa.

The main goal of Free Software is to allow the user to have as much freedom as possible when using 
a piece of software for any purpose. This is in contrast to the traditional alternative, called
Proprietary Software, which can be defined as software that the user can not edit, modify or 
redistribute without the original publishers permission. This kind of software intentionally 
restricts the users freedom, usually for the purpose of profit or control of the software. Some 
common examples of Proprietary Software, are Microsoft's \textit{Windows}, Apple's \textit{iOS}, 
and Google's \textit{Chrome} web browser.

Many people don't know that they already use Free Software\cite{COMMONfoss}, but often the tools
they use most often are Free Software. A few examples of this are, Krita\cite{KRITA}; a graphics 
design and art tool that is used frequently in animation, and other digital art, is made and 
managed by the KDE foundation\cite{KDE}, who make exclusively Free Software. Dovecot\cite{DOVECOT}; 
an email server used by major email providers and is Free Software, A final example is 
Firefox\cite{FIREFOX} a Free Software web browser made by Mozilla that makes up 2.71\% of the 
browser market share as of 2024, however in the past has had up to 30\%\cite{BROWSERmarketshare}.These 
are all more modern examples of Free Software, however over the past 40 years, there have 
been countless others. 

\section{A brief history of FOSS}
The term Free Software was first coined by Richard Stallman in 1983\cite{GNUproject}, however even
before this, examples of Free Software (and the disapproval of Proprietary Software), were already
starting to show. 

One of the earliest examples of the disapproval of Non-free Software, was the response to Microsoft's 
\textit{An open letter to hobbyists}, which was written by Bill Gates in 1976. This letter detailed 
that people had been stealing from Microsoft, as many people had brought hardware through 
them, but far fewer people had brought required software for said hardware. The fact this was happening 
at a scale large enough to cause this showed how many computing groups, also known as hacker groups/spaces, 
weren't willing to pay for the software they used, believing that if they brought the hardware they had done 
all that was needed\cite{OPENletter}. It is often believed that this is one of the first examples 
of \textit{hacker culture}, which would become more common into the 80's and 90's with devices like the blue
box \cite{BLUEbox}.

A key figure in \textit{hacker culture}, as previously mentioned, is Richard Stallman. In the 
 1980's e lheft his job at MIT to work full time on the GNU project, which was designed
to be a full recreation of AT\&T's Unix operating system from the ground up in all Free Software. 
The idea was to allow anyone access to a Unix like machine without paying AT\&T's expensive license 
fees, and allow any user to view it, redistribute or edit; it was to be the first fully free 
operating system. The early development of GNU was relatively slow, and it was not a completely free 
system for many years, as some core parts of the operating system were missing, meaning non-free 
alternatives had to be used. However this would later change in 1991, when final additions would
be created.

In 1988 BSD Net1 would release\cite{BSDnet1}, this was the first fully open version of the Berkeley 
Software Distribution version of Unix. BSD was by no means new by this point, however it wasn't 
fully free until this point It had completely rewritten all the code from the original 
Unix that previous versions contained, meaning it was now completely free from AT\&T's licenses.
It would be the start of a long linage of open source operating systems which are now the base
of MacOS, FreeBSD and OpenBSD.

The GNU project, while still not fully finished, saw the final piece of the puzzle when 
Linux\cite{LINUX} released in 1991, it was a fully open kernel which GNU was still lacking (however 
it did get its own kernel called GNU hurd but Linux is far more commonly used). With GNU and Linux 
paired together a user could finally get a fully free operating system for general use, this 
combination of software is still in use today, having a 4.7\% market share globally on desktop
computers\cite{LINUXmarket}, And on web servers it is dominant. In recent years it has also shown
some use in gaming, with it being the operating system used by Valves \textit{steam deck} gaming 
handheld\cite{STEAMdeck}.

Since Linux's release there haven't been as many major events, however there has been a slow tick
in development, with a large jump over Covid, now Free Software being fully viable against
its Proprietary counterpart.

\section{How is Free Software developed}
The process of developing Free Software has changed over time, especially as the internet came to be,
allowing developers from all across the world to add things. In modern terms the development process is 
very simple, a developer can look at a piece of code, make changes to a local version of it, then it 
can be uploaded to a central online version of the code, to be checked by lead maintainers, before becoming
the part of the main version (developers would say creating a local branch and submitting a pull request). 
This method was popularized version control systems; such as git\cite{GIT}, which is also free software. 
What these tools allow for is the work of many people to brought together into one single code base.

When code is submitted, it generally gets split into individual chunks (called patches) which each
have an individual purpose. Each patch added will fix 1 bug or add 1 feature, this leads to a simple
development cycle that can easily be used to fix bugs, by breaking them down into small patches that
need to be written, and distributing the work between many developers. 

Without going into too much detail, this is done by merging all contributions into the main code base
by comparing line numbers in differing versions, this is a fully automated process, managed by your
version control system. This pattern of development is liked amongst programmers as it allows many 
to submit code all at once, which is invaluable if your project has many developers. This method
is also commonly used in Non-free Software, to manage large development teams\cite{NONFREEvcs}.

\section{Comparing Free Software to its proprietary counterparts}
\section{What makes Free Software so appealing to developers?}
\section{What makes Free Software so appealing to privacy experts?}
\section{Where else is Free Software used and why?}
\section{What's next for the Free Software space?}
\section{Final thoughts}

\newpage
\printbibliography
}
\end{document}
